Reasoning About a Machine with Local Capabilities

### 2018-01-18 esop-2018/paper.tex
Fixed citations through out the paper.

Compromising software security is often based on attacks that break programming
language properties relied upon by software authors, such as control-flow
correctness, local-state encapsulation, etc. Commodity processors offer little
support for defending against such attacks: they offer security primitives with
only coarse-grained memory protection and limited compartmentalization
scalability. As a result, defenses against attacks on control-flow correctness
and local-state encapsulation are either limited to only certain common forms of
attacks (leading to an attack-defense arms race) and/or rely on techniques like
machine code rewriting
(e.g.,~\cite{wahbe_efficient_1993,abadi_control-flow_2005}), machine code
verification (e.g.,~\cite{morrisett_system_1999}), virtual machines with a
native stack (e.g.,~the JVM~\cite{lindholm_java_2014}) or
randomization~\cite{forrest_building_1997}. The latter techniques essentially
emulate protection techniques on existing hardware, at the cost of performance,
system complexity and/or security.

\emph{Capability machines} are a type of processors that
remediate these limitations with a better security model at
the hardware level. They are based on old ideas
(e.g.~\cite{Carter:1994:HSF:195473.195579,Dennis:1966:PSM:365230.365252,shapiro_eros:_1999}),
but have recently received renewed interest; in particular, the CHERI
project has proposed new ideas and ways of tackling practical
challenges like backwards compatibility and realistic OS
support~\cite{Watson2015Cheri,Woodruff:2014:CCM:2665671.2665740}. Capability
machines tag every word (in the register file and in memory) to
enforce a strict separation between numbers and capabilities (a kind
of pointers that carry authority). Memory capabilities carry
the authority to read and/or write to a range of memory
locations. There is also a form of \emph{object capabilities}, which represent the
authority to invoke a piece of code without exposing the code's
encapsulated private state (e.g., the M-Machine's enter capabilities or
CHERI's sealed code/data pairs).

@@

In this paper, we make two main contributions: (1) an alternative calling
convention that uses local capabilities to enforce stack frame encapsulation and
well-bracketed control flow, and (2) perhaps more importantly, we adapt and
apply the well-studied techniques of step-indexed Kripke logical relations for
reasoning about code on a representative capability machine with local
capabilities in general and correctness and security of the calling convention
in particular. More specifically, we make the following contributions:
\begin{itemize}
\item We formalize a simple but representative capability machine featuring local
  capabilities and its operational semantics
  (\sectionname~\ref{sec:capab-mach-with}).
\item We define a novel calling convention enforcing control-flow correctness and
  encapsulation of stack frames (\sectionname~\ref{sec:stack-and-return-pointer}). It
  relies solely on local capabilities and does not require OS support (like a
  trusted stack or call/return instructions).
  %It also does not use separate per-component stacks but a single, fairly standard, contiguous stack.
  %domi: don't mention this here. I think some CSF reviewers misunderstood that we were claiming this as a strength of the approach
  It supports higher-order cross-component calls (e.g., cross-component function
  pointers) and can be efficient assuming only one additional piece of processor
  support: an efficient instruction for clearing a range of memory.
\item We present a novel step-indexed Kripke logical relation for reasoning
  about programs on the capability machine. It is an untyped logical relation,
  inspired by previous work on object capabilities \cite{Devriese:2016ObjCap}.
  We prove an analogue of the standard fundamental theorem of logical relations
  --- to the best of our knowledge, our theorem is % by far
  the most general and powerful formulation of the formal guarantees offered by
  a capability machine (a form of capability
  safety~\cite{Devriese:2016ObjCap,Maffeis2010OC}), including the specific
  guarantees offered for local capabilities. It is very general and not tied to
  our calling convention or a specific way of using the system's capabilities.
  We are the first to apply these techniques for reasoning about capability
  machines and we believe they will prove useful for many other purposes than
  our calling convention.
\item We introduce two novel technical ideas in the unary, step-indexed Kripke
  logical relation used to formulate the above theorem: the use of a
  \emph{single} orthogonal closure (rather than the earlier used biorthogonal
  closure) and a variant of Dreyer et. al.'s public and private future
  worlds~\cite{Dreyer:jfp12} to express the special nature of local
  capabilities. The logical relation and the fundamental theorem expressing
  capability safety are presented in \sectionname~\ref{sec:logical-relation}.
\item We demonstrate our results by applying them to challenging examples,
  specifically constructed to demonstrate local-state encapsulation and
  control-flow correctness guarantees in the presence of cross-component
  function pointers (\sectionname~\ref{sec:examples}). The examples demonstrate both
  the power of our formulation of capability safety and our calling convention.
\end{itemize}
%\lau{I have left out the discussion/future work, related work and conclussion sections from the outline given with the contribution. Do you feel like there should be a dedicated outline instead and if not should we mention the last sections (I don't feel it contributes anything to mention them).}

@@

For technical reasons, we do not actually remove a revoked temporary region from
the world, but we turn it into a special revoked region that exists for this purpose.
Such a revoked region contains no state transition system and puts no
requirements on the memory. It simply serves as a mask for a revoked temporary
region. Masking a region like this goes back to earlier work of
\cite{Ahmed2004semantics} and was also used by
\cite{Thamsborg:2011:KLR:2034773.2034831}.

@@

Regions are used to define safe memory segments, but this set may itself be world-dependent. In other
words, our worlds are defined recursively. Recursive worlds are common
in Kripke models and the following lemma uses the method of
\cite{Birkedal:2011:SKM:1926385.1926401,Birkedal:tutorial-notes} for
constructing them. The formulation of the lemma is technical, so we recommend
that non-expert readers ignore the technicalities and accept that there exists a
set of worlds $\Wor$ and two relations $\futurestr$ and $\futurewk$ satisfying
the (recursive) equations in the theorem (where the $\blater$ operator can be
safely ignored).
% Short technical explanation
% The recursiveness of the worlds gives rise to a recurisve domain
% equation. Theorem~\ref{thm:world-existence} gives the solution to this
% recursive domain equation.
% \dominique{it is not explained what cofes are?}
% \lau{I added what cofe means but no formal definition (I think this goes under the category technicalities)}
\begin{theorem}\label{thm:world-existence}
  There exists a \cofe{} (complete ordered family of equivalences) $\Wor$ and preorders $\futurestr$ and
  $\futurewk$ such that $(\Wor,\futurestr)$ and $(\Wor,\futurewk)$ are
  preordered \cofes{}, and there exists an isomorphism $\xi$ such that
  \begin{align*}
      \xi \; : \; & \Wor \cong \blater (\nats \finparfun \Regions)\\
      \Regions  = & \; \{\revoked\} \uplus \\
               & \; \{\temp\} \times \States \times \Rels \times (\States \fun (\Wor \monwknefun \UPred{\HeapSegments}))\uplus \\
               & \; \{\perma\} \times \States \times \Rels \times (\States \fun (\Wor \monstrnefun \UPred{\HeapSegments}))
    \end{align*}
  and for $W, W' \in \Wor\ldotp \quad  
  \begin{aligned}
    W' \futurestr W & \Leftrightarrow \xi(W') \futurestr \xi(W)   \\
    W' \futurewk W & \Leftrightarrow \xi(W') \futurewk \xi(W)
  \end{aligned}$
\end{theorem}
In the above theorem, $\States \times \Rels$ corresponds to the aforementioned
state transition system where $\Rels$ contains pairs of relations corresponding
to the public and private transitions, and $\States$ is an unspecified set that
we assume to contain at least the states we use in this paper. The last part of
the temporary and permanent regions is a state interpretation function that
determines what memory segments the region permits in each state of the state
transition system.  The different monotonicity requirements in the two
interpretation functions reflects how permanent regions rely only on permanent
protocols whereas temporary regions can rely on both temporary and permanent
protocols.  $\UPred{\HeapSegments}$ is the set of step-indexed, downwards closed
predicates on memory segments:
$\UPred{\HeapSegments} = \{ A \subseteq  \nats \times
\HeapSegments \mid 
\forall (n, ms) \in A. \forall m
\leq n. (m, ms) \in A\}$. 
\

@@

First, the \emph{observation relation} $\observations$ defines what
configurations we consider safe. A configuration is safe
with respect to a world, when the execution of said configuration does not break
the memory protocols of the world. Roughly speaking, this means that when the
execution of a configuration halts, then there is a private future world that
the resulting memory satisfies. Notice that failing is considered safe behavior.
In fact, the machine often resorts to failing when an unauthorized access is
attempted, like loading from a capability without read permission. This is
similar to \cite{Devriese:2016ObjCap}'s logical relation for an untyped
language, but unlike typical logical relations for typed languages, which
require that programs do not fail.

@@

\section{Discussion}
\label{sec:discussion}
\noindent\textbf{Calling convention}\\
\emph{Formulating control flow correctness} While we claim that our calling
convention enforces control-flow correctness, we do not prove a general theorem
that shows this, because it is not clear what such a theorem should look like.
Formulations in terms of a control-flow graph, like the one by
\cite{abadi_control-flow_2005}, do not take into account temporal properties,
like the well-bracketedness that Example~\texttt{\footnotesize{g1}} relies on.
In fact, our examples show that our logical relation imply a stronger form of
control-flow correctness than such formulations, although this is not made very
explicit. As future work, we consider looking at a more explicit and useful way
to formalize control-flow correctness. The idea would be to define a variant
of our capability machine with call and return instructions and
well-bracketed control flow built-in to the operational semantics, and then
prove that compiling such code to our machine using our calling convention is
fully abstract~\cite{abadi_protection_1998}.

@@

Interestingly, \cite{Hur:2011:KLR:1926385.1926402} also use a step-indexed,
Kripke logical relation for an assembly language (for reasoning about correct
compilation from ML to assembly), but because they only model non-adversarial
code that treats return pointers according to a particular calling convention,
they can use standard biorthogonality rather than a single orthogonal closure
like us.

\emph{Public/private future worlds} A novel aspect of our logical relation is
how we model the temporary, revokable nature of local capabilities using public/private
future worlds. The main insight is that this special nature generalizes that of the syntactically-enforced
unstorable status of evaluation contexts in lambda calculi without control
effects (of which well-bracketed control flow is a consequence). To reason about
code that relies on this (particularly, the original awkward example),
\cite{Dreyer:jfp12} (DNB) formally capture the special status of evaluation
contexts using Kripke worlds with public and private future world relations.
Essentially, they allow relatedness of evaluation contexts to be monotone with
respect to a weaker future world relation (public) than relatedness of values,
formalizing the idea that it is safe to make temporary internal state
modifications (private world transitions, which invalidate the continuation,
but not other values) while an expression is performing internal steps, as long
as the code returns to a stable state (i.e. transitions to a public future world
of the original) before returning. We
generalize this idea to reason about local capabilities: validity of local
capabilities is allowed to be monotone with respect to a weaker future-world
relation than other values, which we can exploit to distinguish between state
changes that are always safe (public future worlds) and changes that are only
valid if we clear all local capabilities (private future worlds). Our future
world relations are similar to DNB's (for example, our proof of
the awkward example uses exactly the same state transition system),
but they turn up in an entirely different place in the logical relation: rather
than using public future worlds for the special syntactic category of evaluation
contexts, they are used in the value relation depending on the
locality of the capability at hand. Additionally, our worlds are a bit more
complex because, to allow local memory capabilities and write-local
capabilities, they can contain (revokable) temporary regions that are only
monotonous w.r.t. public future worlds, while DNB's worlds are entirely
permanent.

\emph{Local capabilities in high-level languages} We point out that
local capabilities are quite similar to a feature proposed for the high-level
language Scala: \cite{osvald_gentrification_2016}'s second-class or local
values. They are a kind of values that can be provided to other code for
immediate use without allowing them to be stored in a closure or reference for
later use. We believe reasoning about such values will require techniques similar 
to what we provide for local capabilities.

@@

Capability machines originate with \cite{Dennis:1966:PSM:365230.365252} and we refer to
\cite{Levy1984capability} and \cite{Watson2015Cheri} for an overview
of previous work. The capability machine formalized in
\sectionname~\ref{sec:capab-mach-with} is a simple but representative
model, modeled mainly after the
M-Machine~\cite{Carter:1994:HSF:195473.195579} (the enter pointers
resemble the M-Machine's) and
CHERI~\cite{Watson2015Cheri,Woodruff:2014:CCM:2665671.2665740} (the
memory and local capabilities resemble CHERI's). The latter is a
recent and relatively mature capability machine, which combines
capabilities with a virtual memory approach, in the interest of
backwards compatibility and gradual adoption. As discussed, our local
capabilities can cross module boundaries, 
contrary to what is enforced by CHERI's default CCall implementation.

@@

Our logical relation is a unary, step-indexed Kripke logical relation with
recursive worlds
\cite{pitts_operational_1998,Appel:2001:IMR:504709.504712,Ahmed2004semantics,Birkedal:2011:SKM:1926385.1926401},
closely related to the one used by \cite{Devriese:2016ObjCap} to formulate
capability safety in a high-level JavaScript-like lambda calculus. Our
Fundamental Theorem is similar to theirs and expresses capability safety of the
capability machine. Because we are not interested in externally observable
side-effects (like console output or memory access traces), we do not require
their notion of effect parametricity. Our logical relation uses several ideas
from previous work, like Kripke worlds with regions containing state transition
systems \cite{Ahmed:popl09}, public/private future worlds \cite{Dreyer:jfp12}
(see \sectionname~\ref{sec:discussion} for a discussion), and biorthogonality
\cite{pitts_operational_1998,benton_biorthogonality_2009-1,Hur:2011:KLR:1926385.1926402}.

Swasey et. al.~\cite{swasey:2017} have recently developed a \emph{logic}, OCPL, for verification
of object capability patterns. The logic is based on
Iris~\cite{iris,iris2,iris3}, a state of the art higher-order concurrent
separation logic and is formalized in Coq, building on the Iris Proof Mode for
Coq~\cite{ipm}. OCPL gives a more abstract and modular way of proving
capability safety for a lambda-calculus (with concurrency) compared to the
earlier work by \cite{Devriese:2016ObjCap}.
%domi: rest of the paragraph could be cut?
% In the future we would also like to
% investigate a new program logic for reasoning about capability safety for our
% capability machine model. We think Iris would also be a natural starting point
% for such an endeavour, since Iris is really a framework, which can be
% instantiated to different programming languages. OCPL was able to leverage
% existing Iris specifications for the high-level language; for our capability
% machine model, however, it would be necessary to devise new kinds of
% specifications for our low-level programs with unstructured control-flow. It is
% likely that we could get inspiration from earlier work on logics for assembly
% programming languages, such as XCAP~\cite{xcap}.

>>>>

Compromising software security is often based on attacks that break programming
language properties relied upon by software authors, such as control-flow
correctness, local-state encapsulation, etc. Commodity processors offer little
support for defending against such attacks: they offer security primitives with
only coarse-grained memory protection and limited compartmentalization
scalability. As a result, defenses against attacks on control-flow correctness
and local-state encapsulation are either limited to only certain common forms of
attacks (leading to an attack-defense arms race) and/or rely on techniques like
machine code rewriting~\cite{wahbe_efficient_1993,abadi_control-flow_2005}, machine code
verification~\cite{morrisett_system_1999}, virtual machines with a
native stack~\cite{lindholm_java_2014} or
randomization~\cite{forrest_building_1997}. The latter techniques essentially
emulate protection techniques on existing hardware, at the cost of performance,
system complexity and/or security.

\emph{Capability machines} are a type of processors that
remediate these limitations with a better security model at
the hardware level. They are based on old ideas~\cite{Carter:1994:HSF:195473.195579,Dennis:1966:PSM:365230.365252,shapiro_eros:_1999},
but have recently received renewed interest; in particular, the CHERI
project has proposed new ideas and ways of tackling practical
challenges like backwards compatibility and realistic OS
support~\cite{Watson2015Cheri,Woodruff:2014:CCM:2665671.2665740}. Capability
machines tag every word (in the register file and in memory) to
enforce a strict separation between numbers and capabilities (a kind
of pointers that carry authority). Memory capabilities carry
the authority to read and/or write to a range of memory
locations. There is also a form of \emph{object capabilities}, which represent the
authority to invoke a piece of code without exposing the code's
encapsulated private state (e.g., the M-Machine's enter capabilities or
CHERI's sealed code/data pairs).

@@

In this paper, we make two main contributions: (1) an alternative calling
convention that uses local capabilities to enforce stack frame encapsulation and
well-bracketed control flow, and (2) perhaps more importantly, we adapt and
apply the well-studied techniques of step-indexed Kripke logical relations for
reasoning about code on a representative capability machine with local
capabilities in general and correctness and security of the calling convention
in particular. More specifically, we make the following contributions:
\begin{itemize}
\item We formalize a simple but representative capability machine featuring local
  capabilities and its operational semantics
  (\sectionname~\ref{sec:capab-mach-with}).
\item We define a novel calling convention enforcing control-flow correctness and
  encapsulation of stack frames (\sectionname~\ref{sec:stack-and-return-pointer}). It
  relies solely on local capabilities and does not require OS support (like a
  trusted stack or call/return instructions).
  %It also does not use separate per-component stacks but a single, fairly standard, contiguous stack.
  %domi: don't mention this here. I think some CSF reviewers misunderstood that we were claiming this as a strength of the approach
  It supports higher-order cross-component calls (e.g., cross-component function
  pointers) and can be efficient assuming only one additional piece of processor
  support: an efficient instruction for clearing a range of memory.
\item We present a novel step-indexed Kripke logical relation for reasoning
  about programs on the capability machine. It is an untyped logical relation,
  inspired by previous work on object capabilities \cite{Devriese:2016ObjCap}.
  We prove an analogue of the standard fundamental theorem of logical relations
  --- to the best of our knowledge, our theorem is % by far
  the most general and powerful formulation of the formal guarantees offered by
  a capability machine (a form of capability
  safety~\cite{Devriese:2016ObjCap,Maffeis2010OC}), including the specific
  guarantees offered for local capabilities. It is very general and not tied to
  our calling convention or a specific way of using the system's capabilities.
  We are the first to apply these techniques for reasoning about capability
  machines and we believe they will prove useful for many other purposes than
  our calling convention.
\item We introduce two novel technical ideas in the unary, step-indexed Kripke
  logical relation used to formulate the above theorem: the use of a
  \emph{single} orthogonal closure (rather than the earlier used biorthogonal
  closure) and a variant of Dreyer et al.~\cite{Dreyer:jfp12}'s public and private future
  worlds~\cite{Dreyer:jfp12} to express the special nature of local
  capabilities. The logical relation and the fundamental theorem expressing
  capability safety are presented in \sectionname~\ref{sec:logical-relation}.
\item We demonstrate our results by applying them to challenging examples,
  specifically constructed to demonstrate local-state encapsulation and
  control-flow correctness guarantees in the presence of cross-component
  function pointers (\sectionname~\ref{sec:examples}). The examples demonstrate both
  the power of our formulation of capability safety and our calling convention.
\end{itemize}
%\lau{I have left out the discussion/future work, related work and conclussion sections from the outline given with the contribution. Do you feel like there should be a dedicated outline instead and if not should we mention the last sections (I don't feel it contributes anything to mention them).}

@@

For technical reasons, we do not actually remove a revoked temporary region from
the world, but we turn it into a special revoked region that exists for this purpose.
Such a revoked region contains no state transition system and puts no
requirements on the memory. It simply serves as a mask for a revoked temporary
region. Masking a region like this goes back to earlier work of Ahmed
\cite{Ahmed2004semantics} and was also used by Birkedal et al.~\cite{Thamsborg:2011:KLR:2034773.2034831}.

@@

Regions are used to define safe memory segments, but this set may itself be world-dependent. In other
words, our worlds are defined recursively. Recursive worlds are common
in Kripke models and the following lemma uses the method of
Birkedal and Bizjak~\cite{Birkedal:2011:SKM:1926385.1926401}; Birkedal et al.~\cite{Birkedal:tutorial-notes} for
constructing them. The formulation of the lemma is technical, so we recommend
that non-expert readers ignore the technicalities and accept that there exists a
set of worlds $\Wor$ and two relations $\futurestr$ and $\futurewk$ satisfying
the (recursive) equations in the theorem (where the $\blater$ operator can be
safely ignored).
% Short technical explanation
% The recursiveness of the worlds gives rise to a recurisve domain
% equation. Theorem~\ref{thm:world-existence} gives the solution to this
% recursive domain equation.
% \dominique{it is not explained what cofes are?}
% \lau{I added what cofe means but no formal definition (I think this goes under the category technicalities)}
\begin{theorem}\label{thm:world-existence}
  There exists a \cofe{} (complete ordered family of equivalences) $\Wor$ and preorders $\futurestr$ and
  $\futurewk$ such that $(\Wor,\futurestr)$ and $(\Wor,\futurewk)$ are
  preordered \cofes{}, and there exists an isomorphism $\xi$ such that
  \begin{align*}
      \xi \; : \; & \Wor \cong \blater (\nats \finparfun \Regions)\\
      \Regions  = & \; \{\revoked\} \uplus \\
               & \; \{\temp\} \times \States \times \Rels \times (\States \fun (\Wor \monwknefun \UPred{\HeapSegments}))\uplus \\
               & \; \{\perma\} \times \States \times \Rels \times (\States \fun (\Wor \monstrnefun \UPred{\HeapSegments}))
    \end{align*}
  and for $W, W' \in \Wor\ldotp \quad  
  \begin{aligned}
    W' \futurestr W & \Leftrightarrow \xi(W') \futurestr \xi(W)   \\
    W' \futurewk W & \Leftrightarrow \xi(W') \futurewk \xi(W)
  \end{aligned}$
\end{theorem}
In the above theorem, $\States \times \Rels$ corresponds to the aforementioned
state transition system where $\Rels$ contains pairs of relations corresponding
to the public and private transitions, and $\States$ is an unspecified set that
we assume to contain at least the states we use in this paper. The last part of
the temporary and permanent regions is a state interpretation function that
determines what memory segments the region permits in each state of the state
transition system.  The different monotonicity requirements in the two
interpretation functions reflects how permanent regions rely only on permanent
protocols whereas temporary regions can rely on both temporary and permanent
protocols.  $\UPred{\HeapSegments}$ is the set of step-indexed, downwards closed
predicates on memory segments:
$\UPred{\HeapSegments} = \{ A \subseteq  \nats \times
\HeapSegments \mid 
\forall (n, ms) \in A. \forall m
\leq n. (m, ms) \in A\}$. 
\

@@

First, the \emph{observation relation} $\observations$ defines what
configurations we consider safe. A configuration is safe
with respect to a world, when the execution of said configuration does not break
the memory protocols of the world. Roughly speaking, this means that when the
execution of a configuration halts, then there is a private future world that
the resulting memory satisfies. Notice that failing is considered safe behavior.
In fact, the machine often resorts to failing when an unauthorized access is
attempted, such as loading from a capability without read permission. This is
similar to Devriese et al.~\cite{Devriese:2016ObjCap}'s logical relation for an untyped
language, but unlike typical logical relations for typed languages, which
require that programs do not fail.

@@

\section{Discussion}
\label{sec:discussion}
\noindent\textbf{Calling convention}\\
\emph{Formulating control flow correctness} While we claim that our calling
convention enforces control-flow correctness, we do not prove a general theorem
that shows this, because it is not clear what such a theorem should look like.
Formulations in terms of a control-flow graph, like the one by
Abadi et al.~\cite{abadi_control-flow_2005}, do not take into account temporal properties,
like the well-bracketedness that Example~\texttt{\footnotesize{g1}} relies on.
In fact, our examples show that our logical relation imply a stronger form of
control-flow correctness than such formulations, although this is not made very
explicit. As future work, we consider looking at a more explicit and useful way
to formalize control-flow correctness. The idea would be to define a variant
of our capability machine with call and return instructions and
well-bracketed control flow built-in to the operational semantics, and then
prove that compiling such code to our machine using our calling convention is
fully abstract~\cite{abadi_protection_1998}.

@@

Interestingly, Hur and Dreyer~\cite{Hur:2011:KLR:1926385.1926402} also use a step-indexed,
Kripke logical relation for an assembly language (for reasoning about correct
compilation from ML to assembly), but because they only model non-adversarial
code that treats return pointers according to a particular calling convention,
they can use standard biorthogonality rather than a single orthogonal closure
like us.

\emph{Public/private future worlds} A novel aspect of our logical relation is
how we model the temporary, revokable nature of local capabilities using public/private
future worlds. The main insight is that this special nature generalizes that of the syntactically-enforced
unstorable status of evaluation contexts in lambda calculi without control
effects (of which well-bracketed control flow is a consequence). To reason about
code that relies on this (particularly, the original awkward example),
Dreyer et al.~\cite{Dreyer:jfp12} (DNB) formally capture the special status of evaluation
contexts using Kripke worlds with public and private future world relations.
Essentially, they allow relatedness of evaluation contexts to be monotone with
respect to a weaker future world relation (public) than relatedness of values,
formalizing the idea that it is safe to make temporary internal state
modifications (private world transitions, which invalidate the continuation,
but not other values) while an expression is performing internal steps, as long
as the code returns to a stable state (i.e. transitions to a public future world
of the original) before returning. We
generalize this idea to reason about local capabilities: validity of local
capabilities is allowed to be monotone with respect to a weaker future-world
relation than other values, which we can exploit to distinguish between state
changes that are always safe (public future worlds) and changes that are only
valid if we clear all local capabilities (private future worlds). Our future
world relations are similar to DNB's (for example, our proof of
the awkward example uses exactly the same state transition system),
but they turn up in an entirely different place in the logical relation: rather
than using public future worlds for the special syntactic category of evaluation
contexts, they are used in the value relation depending on the
locality of the capability at hand. Additionally, our worlds are a bit more
complex because, to allow local memory capabilities and write-local
capabilities, they can contain (revokable) temporary regions that are only
monotonous w.r.t. public future worlds, while DNB's worlds are entirely
permanent.

\emph{Local capabilities in high-level languages} We point out that
local capabilities are quite similar to a feature proposed for the high-level
language Scala: Osvald et al.~\cite{osvald_gentrification_2016}'s second-class or local
values. They are a kind of values that can be provided to other code for
immediate use without allowing them to be stored in a closure or reference for
later use. We believe reasoning about such values will require techniques similar 
to what we provide for local capabilities.

@@

Capability machines originate with Dennis and Van Horn~\cite{Dennis:1966:PSM:365230.365252} and we refer to
Levy~\cite{Levy1984capability} and Watson et al.~\cite{Watson2015Cheri} for an overview
of previous work. The capability machine formalized in
\sectionname~\ref{sec:capab-mach-with} is a simple but representative
model, modeled mainly after the
M-Machine~\cite{Carter:1994:HSF:195473.195579} (the enter pointers
resemble the M-Machine's) and
CHERI~\cite{Watson2015Cheri,Woodruff:2014:CCM:2665671.2665740} (the
memory and local capabilities resemble CHERI's). The latter is a
recent and relatively mature capability machine, which combines
capabilities with a virtual memory approach, in the interest of
backwards compatibility and gradual adoption. As discussed, our local
capabilities can cross module boundaries, 
contrary to what is enforced by CHERI's default CCall implementation.

@@

Our logical relation is a unary, step-indexed Kripke logical relation with
recursive worlds
\cite{pitts_operational_1998,Appel:2001:IMR:504709.504712,Ahmed2004semantics,Birkedal:2011:SKM:1926385.1926401},
closely related to the one used by Devriese et al.~\cite{Devriese:2016ObjCap} to formulate
capability safety in a high-level JavaScript-like lambda calculus. Our
Fundamental Theorem is similar to theirs and expresses capability safety of the
capability machine. Because we are not interested in externally observable
side-effects (like console output or memory access traces), we do not require
their notion of effect parametricity. Our logical relation uses several ideas
from previous work, like Kripke worlds with regions containing state transition
systems \cite{Ahmed:popl09}, public/private future worlds \cite{Dreyer:jfp12}
(see \sectionname~\ref{sec:discussion} for a discussion), and biorthogonality
\cite{pitts_operational_1998,benton_biorthogonality_2009-1,Hur:2011:KLR:1926385.1926402}.

Swasey et al.~\cite{swasey:2017} have recently developed a \emph{logic}, OCPL, for verification
of object capability patterns. The logic is based on
Iris~\cite{iris,iris2,iris3}, a state of the art higher-order concurrent
separation logic and is formalized in Coq, building on the Iris Proof Mode for
Coq~\cite{ipm}. OCPL gives a more abstract and modular way of proving
capability safety for a lambda-calculus (with concurrency) compared to the
earlier work by Devriese et al.~\cite{Devriese:2016ObjCap}.
%domi: rest of the paragraph could be cut?
% In the future we would also like to
% investigate a new program logic for reasoning about capability safety for our
% capability machine model. We think Iris would also be a natural starting point
% for such an endeavour, since Iris is really a framework, which can be
% instantiated to different programming languages. OCPL was able to leverage
% existing Iris specifications for the high-level language; for our capability
% machine model, however, it would be necessary to devise new kinds of
% specifications for our low-level programs with unstructured control-flow. It is
% likely that we could get inspiration from earlier work on logics for assembly
% programming languages, such as XCAP~\cite{xcap}.
